A Framework for Substructural Type Systems \thanks{Supported by organization x.}

### 2021-10-15 tex/esop22/algebra.tex
[ writing ] Bunched Typing Rules section
[ writing ] more redrafting near deadline

\begin{definition}
  A \emph{posemiring} is a tuple $(\Ann, \leq, 0, +, 1, *)$ such that the
  following laws hold (universally quantified in all the variables).
  \begin{itemize}
    \item $x \leq x' \to y \leq y' \to x + y \leq x' + y'$
    \item $x \leq x' \to y \leq y' \to x * y \leq x' * y'$
    \item $0 + x = x
      \land x + 0 = x
      \land (x + y) + z = x + (y + z)
      \land x + y = y + x$
    \item $1 * x = x
      \land x * 1 = x
      \land (x * y) * z = x * (y * z)$
    \item $0 * x = 0
      \land (x + y) * z = x * z + y * z$
    \item $x * 0 = 0
      \land x * (y + z) = x * y + x * z$
  \end{itemize}
\end{definition}

@@

\begin{figure}
  \begin{align*}
    \dot1\,\grR &\coloneqq 1 \\
    (T \dottimes U)\,\grR &\coloneqq T\,\grR \times U\,\grR \\
    (T \dotto U)\,\grR &\coloneqq T\,\grR \to U\,\grR \\
    I^*\,\grR &\coloneqq \grR \leq \gr0 \\
    (T \sep U)\,\grR &\coloneqq \Sigma \grP,\grQ.~\plr{\grR \leq \grP + \grQ}
                       \times T\,\grP \times U\,\grQ \\
    (\gr r \cdot T)\,\grR &\coloneqq \Sigma \grP.~\plr{\grR \leq \gr r\grP}
                       \times T\,\grP \\
    (T \wand U)\,\grP &\coloneqq \Pi \grQ,\grR.~\plr{\grR \leq \grP + \grQ}
                       \to T\,\grQ \to U\,\grR
  \end{align*}
  \caption{The bunched connectives}
  \label{fig:bunched}
\end{figure}

>>>>

\begin{definition}
  A \emph{posemiring} is a tuple $(\Ann, \leq, 0, +, 1, *)$ such that
  $(\Ann, \leq)$ is a partially ordered set, $(\Ann, 0, +)$ is a commutative
  monoid, $(\Ann, 1, *)$ is a monoid, $+$ and $*$ are monotonic, and $*$
  distributes over $0$ and $+$ in the usual way on both sides.
  %the
  %following laws hold (universally quantified in all the variables).
  %\begin{itemize}
  %  \item $x \leq x' \to y \leq y' \to x + y \leq x' + y'$
  %  \item $x \leq x' \to y \leq y' \to x * y \leq x' * y'$
  %  \item $0 + x = x
  %    \land x + 0 = x
  %    \land (x + y) + z = x + (y + z)
  %    \land x + y = y + x$
  %  \item $1 * x = x
  %    \land x * 1 = x
  %    \land (x * y) * z = x * (y * z)$
  %  \item $0 * x = 0
  %    \land (x + y) * z = x * z + y * z$
  %  \item $x * 0 = 0
  %    \land x * (y + z) = x * y + x * z$
  %\end{itemize}
\end{definition}

\begin{example}[Zero-one-many]\label{thm:linearity}
  The poset $\{0 > \omega < 1\}$ forms a posemiring under normal numeric
  addition (with $1 + 1 = 1 + \omega = \omega + \omega = \omega$) and
  multiplication (with $\omega * \omega = \omega$).
  This gives us a way to mark whether variables are unused ($0$), used linearly
  ($1$), or used unrestrictedly ($\omega$) in the current (sub)term.
  The ordering says that unrestricted-use variables can also remain unused or
  be used linearly.
\end{example}

\begin{example}[Variance]\label{thm:variance}
  The set
  $\{{\sim\sim}, {\uparrow\uparrow}, {\downarrow\downarrow}, {\wn\wn}\}$,
  with ${\sim\sim}$ at the bottom and ${\wn\wn}$ at the top of the order, forms
  a posemiring with addition being \emph{meet}, $0$ being \emph{top}
  (${\wn\wn}$), $1$ being ${\uparrow\uparrow}$, and multiplication being
  commutative and determined by
  ${\downarrow\downarrow} * {\downarrow\downarrow} = {\uparrow\uparrow}$ and
  ${\sim\sim} * {\downarrow\downarrow} = {\sim\sim} * {\sim\sim} = {\sim\sim}$.
  This gives us a way to track the variance with which variables are used, in
  the aim of all terms being monotonic in their free variables.
  ${\uparrow\uparrow}$ stands for covariance, ${\downarrow\downarrow}$ for
  contravariance, ${\sim\sim}$ for invariance, and ${\wn\wn}$ for a variable
  with no guarantees, in which we must be constant.
\end{example}

@@

### 2021-10-15 tex/esop22/intro.tex
[ writing ] more redrafting near deadline

\begin{itemize}
  \item Review standard linear algebra constructions in the posemiring setting.
  \item Give a general definition of simple type systems with usage annotations
    based on connectives from bunched logic and constructions from linear
    algebra.
  \item Give some novel examples of such systems.
  \item Derive admissibility of renaming and substitution for all such systems.
  \item Give a usage checker generic in syntax.
\end{itemize}

@@

>>>>

%\begin{itemize}
%  \item Review standard linear algebra constructions in the posemiring setting.
%  \item Give a general definition of simple type systems with usage annotations
%    based on connectives from bunched logic and constructions from linear
%    algebra.
%  \item Give some novel examples of such systems.
%  \item Derive admissibility of renaming and substitution for all such systems.
%  \item Give a usage checker generic in syntax.
%\end{itemize}

@@

This paper proceeds as follows.
In \cref{sec:algebra}, we review and fix conventions pertaining to partially
ordered semirings and vectors over them.
In \cref{sec:bunched-rules}, we introduce an informal meta-syntax allowing us
to state substructural typing rules succinctly and without explicit reference
to contexts.
In \cref{sec:syntax}, we mechanise that meta-syntax, giving a type of
\emph{descriptions} of type systems, and interpreting those descriptions as
types of intrinsically typed terms.
In \cref{sec:env}, we discuss usage-aware environments: a generalisation of
the structures used in simultaneous renaming and substitution proofs.
We use environments in \cref{sec:semantics} to state an alternative
elimination principle for terms, and give examples of such eliminations in
\cref{sec:examples}.
The examples are syntax-generic renaming and substitution, a specific
denotational semantics, and a syntax-generic usage elaborator.
Finally, we conclude and discuss future work in \cref{sec:conc}

### 2021-10-15 tex/esop22/syntax.tex
[ writing ] Bunched Typing Rules section
[ writing ] more redrafting near deadline

\begin{figure}
  \begin{displaymath}
    \begin{prooftree}[comb]
      \hypo{\sqni A}
      \infer1[Var]{\vdash A}
    \end{prooftree}
    \qquad
    \begin{prooftree}[comb]
      \hypo{\gr1A \vdash B}
      \infer1[$\multimap$-I]{\vdash A \multimap B}
    \end{prooftree}
    \qquad
    \begin{prooftree}[comb]
      \hypo{\vdash A \multimap B}
      \hypo{\sep}
      \hypo{\vdash A}
      \infer3[$\multimap$-E]{\vdash B}
    \end{prooftree}
  \end{displaymath}
  \begin{displaymath}
    \begin{prooftree}[comb]
      \hypo{\vdash A_i}
      \infer1[$\oplus$-I$_i$]{\vdash A_0 \oplus A_1}
    \end{prooftree}
    \qquad
    \begin{prooftree}[comb]
      \hypo{\vdash A \oplus B}
      \hypo{\sep}
      \hypo{(\gr1A \vdash C}
      \hypo{\dottimes}
      \hypo{\gr1B \vdash C)}
      \infer5[$\oplus$-E]{\vdash C}
    \end{prooftree}
  \end{displaymath}
  \begin{displaymath}
    \begin{prooftree}[comb]
      \hypo{\gr r \cdot \plr{\vdash A}}
      \infer1[$\oc_{\gr r}$-I]{\vdash \oc_{\gr r} A}
    \end{prooftree}
    \qquad
    \begin{prooftree}[comb]
      \hypo{\vdash \oc_{\gr r} A}
      \hypo{\sep}
      \hypo{\gr rA \vdash C}
      \infer3[$\oc_{\gr r}$-E]{\vdash C}
    \end{prooftree}
  \end{displaymath}
  \caption{\cref{fig:lr} restated in terms of bunched combinators}
  \label{fig:lr-comb}
\end{figure}

@@

The next thing we want to do is to build terms in the described type system.

@@

Here is an example term, using the \AgdaFunction{$\lambda$R} system.
First, for ease of writing, we introduce pattern synonyms for each of the
typing rules we use.

\ExecuteMetaData[\PaperExamplestex]{patterns}

Our example term is a function that flips a tagged union wrapped in an
arbitrarily annotated \emph{bang}.
Much of the effort in writing such a term goes into writing the various
relatedness proofs between usage contexts --- observing, for example, that two
usage contexts sum together to make a third, or that a usage context used for
a variable is a basis vector.
We give a method of automating these proofs in \cref{sec:usage-elaborator}.
\todo{To be clear, we don't actually write this.}

\ExecuteMetaData[\HeavyItex]{lR-term}

A layer of syntax supports the following functorial action.
{\color{red}(Maybe this has to go later, because it talks about linear maps.)}

>>>>

@@

The next thing we want to do is to build terms in the described type system.
The following definitions are useful for talking about types indexed over
contexts, judgement forms, and judgement forms admitting newly bound variables,
respectively.

@@

Terms defined like this are still quite difficult to write, mainly because of
frequently changing usage contexts and the need for proofs that they all match
up.
We will see how to automate these proofs in \cref{sec:usage-elaborator}.

%Here is an example term, using the \AgdaFunction{$\lambda$R} system.
%First, for ease of writing, we introduce pattern synonyms for each of the
%typing rules we use.

%\ExecuteMetaData[\PaperExamplestex]{patterns}

%Our example term is a function that flips a tagged union wrapped in an
%arbitrarily annotated \emph{bang}.
%Much of the effort in writing such a term goes into writing the various
%relatedness proofs between usage contexts --- observing, for example, that two
%usage contexts sum together to make a third, or that a usage context used for
%a variable is a basis vector.
%We give a method of automating these proofs in \cref{sec:usage-elaborator}.
%\todo{To be clear, we don't actually write this.}

%\ExecuteMetaData[\HeavyItex]{lR-term}

A layer of syntax supports the following functorial action.
